---
title: "JOINT ANALYSES SLE PI-LowC3 AND OUTCOME"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### JOIN ANALYSES WITH R RANDOM COEFFICIENT - JMBayes2 R package



```{r}
rm( list=ls() )
setwd("C:/Documenti/Manenti/LES/JOHNS HOPKINS")
library(haven)
library(tidyverse)
sle_status_cross_sectional <- read_dta("sle_status_cross_sectional.dta")
sle_05_long <- read_dta("sle_05_long.dta")
sle_05_time_varying <- read_dta("sle_05_time_varying.dta")
```


# JMbayes2 require that all datset are sorted by id
```{r}
sle_status_cross_sectional <- sle_status_cross_sectional %>% 
  mutate(id = as.numeric(id)) %>% 
  arrange(id)
sle_05_long <- sle_05_long %>% 
  mutate(id = as.numeric(id)) %>% 
  arrange(id)
sle_05_time_varying <- sle_05_time_varying %>% 
  mutate(id = as.numeric(id)) %>% 
  arrange(id)
sle_05_long$LOWC3 <- as.factor(sle_05_long$LOWC3)
sle_status_cross_sectional$LOWC3 <- as.factor(sle_status_cross_sectional$LOWC3)
sle_05_time_varying$LOWC3 <- as.factor(sle_05_time_varying$LOWC3)
```

### ANALYSES WITH JMbayes2
```{r }
options(scipen=999)
library(JMbayes2)
```

### ADJUSTED ANALYSES

## ADJUSTED COX PH REGRESSION ANALYSIS WITH TIME-VARYING eGFR
```{r}
Stand.TV.CoxFit <- coxph(Surv(tstart, tstop, status) ~ LOWC3 + eGFR +
	ageatbiopsy + sex +  black + hypertension + mmf + cyc, 
	data = sle_05_time_varying, cluster = id)
summary(Stand.TV.CoxFit)
```


## ADJUSTED LONGITUDINAL ANALYSIS - random slope regression
```{r }
lmeFit.rs.sle <- lme(eGFR ~ LOWC3 + time +
	ageatbiopsy + sex + black + hypertension + mmf + cyc, 
	random = ~ 1 + time | id, control = lmeControl(opt = 'optim'), data = sle_05_long)
summary(lmeFit.rs.sle)
```

## ADJUSTED COX PH REGRESSION ANALYSIS NOT INCLUDING eGFR
```{r }
CoxFit <- coxph(Surv(tstart, tstop, status) ~ LOWC3 + 
	ageatbiopsy + sex +  black + hypertension + mmf + cyc, 
	data = sle_05_time_varying, cluster = id)
summary(CoxFit)
```

## with eGFR current value
```{r }
jointFit <- jm(CoxFit, lmeFit.rs.sle, time_var = "time", n_chains = 4L, n_thin = 50, n_iter = 10000, data_Surv = sle_05_time_varying )
```

## with eGFR slope
```{r }
fForms <- list(
  "eGFR" = ~ slope(eGFR) 
)
jointFit1 <- update(jointFit, functional_forms = fForms)
```

## with eGFR AUC
```{r, message=FALSE, warning = FALSE }
fForms <- list(
     "eGFR" = ~ area(eGFR)
	 )
jointFit2 <- update(jointFit, functional_forms = fForms)
```

## ADJUSTED JOINT LONGITUDINAL SURVIVAL ANALYSIS - current eGFR value
```{r }
summary(jointFit)
```
## ADJUSTED JOINT LONGITUDINAL SURVIVAL ANALYSIS - slope eGFR
```{r }
summary(jointFit1)
```
## ADJUSTED JOINT LONGITUDINAL SURVIVAL ANALYSIS - normalized area/cumulative-effect eGFR
```{r }
summary(jointFit2)
```
## COMPARE MODELS - ADJUSTED
```{r }
compare_jm(jointFit,jointFit1, jointFit2)
```

# DISPLAY COMPACT P values and 95% credible intervals
```{r }
jmb2.ci.p <- function(x, times = 1) {
		b.m <-  mean(c.all)
		b.sd <- sd(c.all)
		b.ci <- quantile(c.all, probs = c(0.025, 0.975))
		tw <- c(mean(c.all <= 0), mean(c.all >= 0))
		prob.not = min(tw)
	    p <- 2 * prob.not
		hr <- exp(mean(c.all) * times)
	    ci <- exp(quantile(c.all, probs = c(0.025, 0.975)) * times)
		if (ci[2] < ci[1]) {
	    cat("HR: ", round(hr,2), "(95%CI: ", round(ci[2],2), " to ", round(ci[1],2), "; P= ", round(p, 4) , ")\n")
		} else {
		cat("HR: ", round(hr,2), "(95%CI: ", round(ci[1],2), " to ", round(ci[2],2), "; P= ", round(p, 4) , ")\n")
		}
	}
```	

## ESTIMATES FROM THE MODELS - current value - adjusted
```{r }
b <- coef(jointFit)
b
```	

# reproducing JMbayes2 output - current value - adjusted
```{r }
c.1 <- jointFit$mcmc[[3]][[1]][,1]
c.2 <- jointFit$mcmc[[3]][[2]][,1]
c.3 <- jointFit$mcmc[[3]][[3]][,1]
c.4 <- jointFit$mcmc[[3]][[4]][,1]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all)
```	

# calculation association parameter (per 10mL/min/1.73m2 decrease) - current value -adjusted
```{r }
c.1 <- jointFit$mcmc[[4]][[1]]
c.2 <- jointFit$mcmc[[4]][[2]]
c.3 <- jointFit$mcmc[[4]][[3]]
c.4 <- jointFit$mcmc[[4]][[4]]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all, times = -10)
```	


## ESTIMATES FROM THE MODELS - slope eGFR - adjusted
```{r }
b <- coef(jointFit1)
b
```	

# calculations: reproducing JMbayes2 output - slope eGFR - adjusted
```{r }
c.1 <- jointFit1$mcmc[[3]][[1]][,1]
c.2 <- jointFit1$mcmc[[3]][[2]][,1]
c.3 <- jointFit1$mcmc[[3]][[3]][,1]
c.4 <- jointFit1$mcmc[[3]][[4]][,1]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all)
```	

# calculation association parameter (per 10mL/min/1.73m2 decrease) - slope eGFR - adjusted
```{r }
c.1 <- jointFit1$mcmc[[4]][[1]]
c.2 <- jointFit1$mcmc[[4]][[2]]
c.3 <- jointFit1$mcmc[[4]][[3]]
c.4 <- jointFit1$mcmc[[4]][[4]]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all, times = 1)
```	

## ESTIMATES FROM THE MODELS - normalized area/cumulative-effect eGFR-adjusted
```{r }
b <- coef(jointFit2)
b
```	

# calculations: reproducing JMbayes2 output - normalized area/cumulative-effect eGFR -adjusted
```{r }
c.1 <- jointFit2$mcmc[[3]][[1]][,1]
c.2 <- jointFit2$mcmc[[3]][[2]][,1]
c.3 <- jointFit2$mcmc[[3]][[3]][,1]
c.4 <- jointFit2$mcmc[[3]][[4]][,1]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all)
```	

# calculation association parameter (per 10mL/min/1.73m2 decrease) - normalized area/cumulative-effect eGFR - adjusted
```{r }
c.1 <- jointFit2$mcmc[[4]][[1]]
c.2 <- jointFit2$mcmc[[4]][[2]]
c.3 <- jointFit2$mcmc[[4]][[3]]
c.4 <- jointFit2$mcmc[[4]][[4]]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all, times = -10)
```	

### CRUDE ANALYSES
## CRUDE LONGITUDINAL ANALYSIS - random slope regression
```{r }
lmeFit.rs.sle <- lme(eGFR ~ LOWC3 + time, 
	random = ~ 1 + time | id, data = sle_05_long)
```	

## CRUDE COX PH REGRESSION ANALYSIS
```{r }
CoxFit <- coxph(Surv(tstart, tstop, status) ~ LOWC3, 
	data = sle_05_time_varying, cluster = id)
```	

## CRUDE COX PH REGRESSION ANALYSIS WITH TIME-VARYING eGFR
```{r }
Stand.TV.CoxFit <- coxph(Surv(tstart, tstop, status) ~ LOWC3 + eGFR, 
	data = sle_05_time_varying, cluster = id)
```	

## ADJUSTED COX PH REGRESSION ANALYSIS NOT INCLUDING eGFR
```{r }
CoxAFit <- coxph(Surv(tstart, tstop, status) ~ LOWC3 + mmf + ageatbiopsy + sex +  black + hypertension, 
	data = sle_05_time_varying, cluster = id)
```	
	
## stardard Cox PH model crude
```{r }
summary(CoxFit)
```	
## standard Cox PH model adjusted
```{r }
summary(CoxAFit)
```	
## with eGFR current value - crude
```{r, message=FALSE, warning = FALSE }
jointCFit <- jm(CoxFit, lmeFit.rs.sle, time_var = "time", n_chains = 4L, n_thin = 50, n_iter = 10000, data_Surv = sle_05_time_varying )
```	

## with eGFR slope - crude
```{r }
fForms <- list(
  "eGFR" = ~ slope(eGFR) 
)
jointCFit1 <- update(jointCFit, functional_forms = fForms)
```	

## with eGFR AUC  - crude
```{r }
fForms <- list(
     "eGFR" = ~ area(eGFR)
	 )
jointCFit2 <- update(jointCFit, functional_forms = fForms)
```	

## CRUDE JOINT LONGITUDINAL SURVIVAL ANALYSIS - current eGFR value
```{r }
summary(jointCFit)
```	
## CRUDE JOINT LONGITUDINAL SURVIVAL ANALYSIS - slope eGFR
```{r }
summary(jointCFit1)
```	
## CRUDE JOINT LONGITUDINAL SURVIVAL ANALYSIS - normalized area/cumulative-effect eGFR
```{r}
summary(jointCFit2)
```	

# DISPLAY COMPACT P values and 95% credible intervals
```{r }
jmb2.ci.p <- function(x, times = 1) {
		b.m <-  mean(c.all)
		b.sd <- sd(c.all)
		b.ci <- quantile(c.all, probs = c(0.025, 0.975))
		tw <- c(mean(c.all <= 0), mean(c.all >= 0))
		prob.not = min(tw)
	    p <- 2 * prob.not
		hr <- exp(mean(c.all) * times)
	    ci <- exp(quantile(c.all, probs = c(0.025, 0.975)) * times)
		if (ci[2] < ci[1]) {
	    cat("HR: ", round(hr,2), "(95%CI: ", round(ci[2],2), " to ", round(ci[1],2), "; P= ", round(p, 4) , ")\n")
		} else {
		cat("HR: ", round(hr,2), "(95%CI: ", round(ci[1],2), " to ", round(ci[2],2), "; P= ", round(p, 4) , ")\n")
		}
	}
```		



## ESTIMATES FROM THE MODELS - current value - crude
```{r }
b <- coef(jointCFit)
b
```

# calculations: reproducing JMbayes2 output - current value - crude
```{r }
c.1 <- jointCFit$mcmc[[3]][[1]][,1]
c.2 <- jointCFit$mcmc[[3]][[2]][,1]
c.3 <- jointCFit$mcmc[[3]][[3]][,1]
c.4 <- jointCFit$mcmc[[3]][[4]][,1]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all)
```

# calculation association parameter (per 10mL/min/1.73m2 decrease) - current value - crude
```{r }
c.1 <- jointCFit$mcmc[[4]][[1]]
c.2 <- jointCFit$mcmc[[4]][[2]]
c.3 <- jointCFit$mcmc[[4]][[3]]
c.4 <- jointCFit$mcmc[[4]][[4]]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all, times = -10)
```


## ESTIMATES FROM THE MODELS - slope eGFR - crude
```{r }
b <- coef(jointCFit1)
b
```

# calculations: reproducing JMbayes2 output - slope eGFR - crude
```{r }
c.1 <- jointCFit1$mcmc[[3]][[1]][,1]
c.2 <- jointCFit1$mcmc[[3]][[2]][,1]
c.3 <- jointCFit1$mcmc[[3]][[3]][,1]
c.4 <- jointCFit1$mcmc[[3]][[4]][,1]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all)
```

# calculation association parameter (per 10mL/min/1.73m2 decrease) - slope eGFR - crude
```{r }
c.1 <- jointCFit1$mcmc[[4]][[1]]
c.2 <- jointCFit1$mcmc[[4]][[2]]
c.3 <- jointCFit1$mcmc[[4]][[3]]
c.4 <- jointCFit1$mcmc[[4]][[4]]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all, times = 1)
```

## ESTIMATES FROM THE MODELS - normalized area/cumulative-effect eGFR - crude
```{r }
b <- coef(jointCFit2)
b
```


#calculations: reproducing JMbayes2 output - normalized area/cumulative-effect eGFR
```{r }
c.1 <- jointCFit2$mcmc[[3]][[1]][,1]
c.2 <- jointCFit2$mcmc[[3]][[2]][,1]
c.3 <- jointCFit2$mcmc[[3]][[3]][,1]
c.4 <- jointCFit2$mcmc[[3]][[4]][,1]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all)
```

# calculation association parameter (per 10mL/min/1.73m2 decrease) - normalized area/cumulative-effect eGFR - crude
```{r }
c.1 <- jointCFit2$mcmc[[4]][[1]]
c.2 <- jointCFit2$mcmc[[4]][[2]]
c.3 <- jointCFit2$mcmc[[4]][[3]]
c.4 <- jointCFit2$mcmc[[4]][[4]]
c.all <- c(c.1,c.2,c.3,c.4)
jmb2.ci.p(c.all, times = -10)
```


## COMPARE MODELS - ADJUSTED AND CRUDE
```{r }
compare_jm(jointFit,jointFit1, jointFit2)
compare_jm(jointCFit,jointCFit1, jointCFit2)
```

## Compare parameters of the longitudinal model
```{r }
long.beta <-    fixef(jointFit)
long.beta1 <-   fixef(jointFit1)
long.beta2 <-   fixef(jointFit2)
long.Cbeta <-   fixef(jointCFit)
long.Cbeta1 <-  fixef(jointCFit1)
long.Cbeta2 <-  fixef(jointCFit2)
long.beta
long.beta1
long.beta2
long.Cbeta
long.Cbeta1
long.Cbeta2
```


## random effects survival model
```{r }
re.surv <- ranef(jointFit)[,1]
re.surv1 <- ranef(jointFit1)[,1]
re.surv2 <- ranef(jointFit2)[,1]
re.Csurv <- ranef(jointCFit)[,1]
re.Csurv1 <- ranef(jointCFit1)[,1]
re.Csurv2 <- ranef(jointCFit2)[,1]
```

## random effects longitudinal model
```{r }
re.long <- ranef(jointFit)[,2]
re.long1 <- ranef(jointFit1)[,2]
re.long2 <- ranef(jointFit2)[,2]
re.Clong <- ranef(jointCFit)[,2]
re.Clong1 <- ranef(jointCFit1)[,2]
re.Clong2 <- ranef(jointCFit2)[,2]
```

### traceplots (scroll with left and right arrow)
```{r }
traceplt <- ggtraceplot(jointFit)
traceplt1 <- ggtraceplot(jointFit1)
traceplt2 <- ggtraceplot(jointFit2)
```
## ADJUSTED
```{r }
traceplt
traceplt1
traceplt2
```
##CRUDE
```{r }
traceCplt <- ggtraceplot(jointCFit)
traceCplt1 <- ggtraceplot(jointCFit1)
traceCplt2 <- ggtraceplot(jointCFit2)
traceCplt
traceCplt1
traceCplt2
```

### density plots (scroll with left and right arrow)
```{r }
densplt <- ggdensityplot(jointFit)
densplt1 <- ggdensityplot(jointFit1)
densplt2 <- ggdensityplot(jointFit2)
```
## ADJUSTED
```{r }
densplt
densplt1
densplt2
```
## CRUDE
```{r }
densCplt <- ggdensityplot(jointCFit)
densCplt1 <- ggdensityplot(jointCFit1)
densCplt2 <- ggdensityplot(jointCFit2)
densCplt
densCplt1
densCplt2
```
# Save analyses
```{r }
name <- paste("C:/Documenti/Manenti/LES/JOHNS HOPKINS/jm_analysis_sle_lowc3", format(as.Date(Sys.Date() - 1), "%d.%m.%Y"), ".RData", sep = "")
save.image(name)
```
